---
title: "HAR_contact_fetal_atlas_expression_bias"
output: html_document
date: "2023-09-21"
---

Load in necessary packages

```{r setup}
require(Seurat)
require(SeuratDisk)
require(tidyverse)
require(ComplexHeatmap)
require(Nebulosa)
require(scales)
require(forcats)
```

Generating heatmaps of clustered average scaled expression values for NSC HAR contacts

``` {r create_NSC_heatmaps}
# Load in fetal and embryonic human progenitor atlas (Bhaduri et al. 2021, Eze et al. 2021)
fetal_human_atlas_seurat <- LoadH5Seurat(file = "fetal_human_atlas_seurat_progs.h5Seurat")

# Load in HAR contacts from in vitro NSCs 
NSC_HAR_targets <- read.table("./contact_lists/mapping_human.txt")
colnames(NSC_HAR_targets) <- c("HAR", "target")
NSC_HAR_targets <- separate_rows(NSC_HAR_targets, target, sep=",")

# Scale data and extract average scaled expression values for all NSC HAR contacts
fetal_human_atlas_seurat_progs <- ScaleData(fetal_human_atlas_seurat_progs)
NSC_contacts_atlas_avg_exp <- AverageExpression(fetal_human_atlas_seurat_progs,
                                                     features = unique(NSC_HAR_targets$target), slot = "scale.data")
NSC_contacts_atlas_avg_exp <- NSC_contacts_atlas_avg_exp$RNA

# Extract average normalized expression values for all NSC HAR contacts (for downstream ordering of heatmaps)
NSC_contacts_atlas_avg_norm_exp <- AverageExpression(fetal_human_atlas_seurat_progs,
                                                     features = unique(NSC_HAR_targets$target), slot = "data")
NSC_contacts_atlas_avg_norm_exp <- NSC_contacts_atlas_avg_norm_exp$RNA

# Alter names of progenitor subtypes for downstream visualization
colnames(NSC_contacts_atlas_avg_exp) <- c("CB_NSC2 (G1/S)", "CB_NSC2 (G2/M)", "CB_NSC (G1/S)", "CB_NSC (G2/M)", "CB_NSC3 (G1/S)", "CGE_NSC (G1/S)", "CGE_NSC (G2/M)", "CX_IPC (G1/S)", "CX_IPC (G2/M)", "CX_ORG (G1/S)", "CX_RG (G1/S)", "CX_RG (G2/M)", "HC_NSC2 (G1/S)", "HC_NSC2 (G2/M)", "HC_NSC (G1/S)", "HC_NSC (G2/M)", "HC_NSC3 (G1/S)", "HT_NSC2 (G1/S)", "HT_NSC2 (G2/M)", "HT_NSC (G1/S)", "HT_NSC (G2/M)", "HT_NSC3 (G1/S)", "HT_NSC3 (G2/M)", "LGE_NSC (G1/S)", "LGE_NSC (G2/M)", "MGE_NSC (G1/S)", "MGE_NSC (G2/M)", "TH_NSC (G1/S)", "TH_NSC (G2/M)")
colnames(NSC_contacts_atlas_avg_exp) <- str_split_fixed(colnames(NSC_contacts_atlas_avg_exp), " ", 2)[,1]
colnames(NSC_contacts_atlas_avg_norm_exp) <- c("CB_NSC2 (G1/S)", "CB_NSC2 (G2/M)", "CB_NSC (G1/S)", "CB_NSC (G2/M)", "CB_NSC3 (G1/S)", "CGE_NSC (G1/S)", "CGE_NSC (G2/M)", "CX_IPC (G1/S)", "CX_IPC (G2/M)", "CX_ORG (G1/S)", "CX_RG (G1/S)", "CX_RG (G2/M)", "HC_NSC2 (G1/S)", "HC_NSC2 (G2/M)", "HC_NSC (G1/S)", "HC_NSC (G2/M)", "HC_NSC3 (G1/S)", "HT_NSC2 (G1/S)", "HT_NSC2 (G2/M)", "HT_NSC (G1/S)", "HT_NSC (G2/M)", "HT_NSC3 (G1/S)", "HT_NSC3 (G2/M)", "LGE_NSC (G1/S)", "LGE_NSC (G2/M)", "MGE_NSC (G1/S)", "MGE_NSC (G2/M)", "TH_NSC (G1/S)", "TH_NSC (G2/M)")
colnames(NSC_contacts_atlas_avg_norm_exp) <- str_split_fixed(colnames(NSC_contacts_atlas_avg_norm_exp), " ", 2)[,1]

# Group the average scaled expression values of NSC HAR contacts into 15 clusters using kmeans
num_centers = 15
NSC_contacts_atlas_kmeans <- kmeans(NSC_contacts_atlas_avg_exp, centers = num_centers)

# Establish color scheme for eight brain regions and cell cycle phase
cell_cycle_colors <- c("light grey", "#386C0B")
names(cell_cycle_colors) <- c("(G1/S)", "(G2/M)")
region_colors <- hue_pal()(8)
names(region_colors) <- sort(unique(fetal_human_atlas_seurat_progs$origin))

# Extract regional origin and cell cycle from atlas metadata and associate with respective color map
regional_origin <- fetal_human_atlas_seurat_progs@meta.data %>% select(cell_type, origin) %>% group_by(cell_type) %>% summarize(regional_origin = names(which.max(table(origin)))) %>% mutate(regional_color = region_colors[regional_origin]) %>% mutate(cell_cycle = str_split_fixed(cell_type, " ", 2)[,2])
regional_origin$cell_cycle[5] <- "(G1/S)"
regional_origin$cell_cycle[17] <- "(G1/S)"
regional_origin$cell_cycle_colors <- cell_cycle_colors[regional_origin$cell_cycle]

# Create regional origin and cell cycle phase annotations for heatmaps
region_annotation <- HeatmapAnnotation(region = regional_origin$regional_origin, col = list("region" = regional_origin$regional_color), annotation_label = "Regional Origin", annotation_name_gp = gpar(fontsize = 8), show_legend = FALSE)
cell_cycle_annotation <- HeatmapAnnotation(cell_cycle = regional_origin$cell_cycle, col = list("cell_cycle" = regional_origin$cell_cycle_colors), annotation_label = "Cell Cycle Phase", annotation_name_gp = gpar(fontsize = 8), show_legend = FALSE)

# Iterate through each cluster, creating a heatmap visualizing average scaled expression for genes within that cluster, ordered by the greatest normalized expression of each gene across NSC subtypes
ann_heatmaps_by_cluster <- lapply(1:num_centers, function(x) {
draw(Heatmap(as.matrix(NSC_contacts_atlas_avg_exp[NSC_contacts_atlas_kmeans$cluster == x,][names(sort.int(apply(NSC_contacts_atlas_avg_norm_exp[rownames(NSC_contacts_atlas_avg_exp[NSC_contacts_atlas_kmeans$cluster == x,]),], 1, max), decreasing = TRUE)),]),
                                          name = glue::glue("Cluster { x }"),
                                          bottom_annotation = cell_cycle_annotation,
                                          top_annotation = region_annotation,
                                          row_names_gp = gpar(fontsize = 10),
                                          show_column_dend = TRUE,
                                          show_row_dend = FALSE,
                                          cluster_columns = TRUE,
                                          cluster_rows = FALSE,
                                          show_heatmap_legend = FALSE,
                                          column_names_rot = 70), padding = unit(c(2, 10, 2, 2), "mm"))})

# Annotation of heatmaps and average expression by NSC subtype or regional bias
names(ann_heatmaps_by_cluster) <- c("Low Cerebellum", "Hippocampus/Cortex", "Cortical oRG", "G2/M", "Hypothalamus", "Cortical IPC", "Ventral", "oRG", "Hippocampus/Cortex", "Low Bias", "Thalamus/Cerebellum",  "Cerebellum", "Mostly G2/M", "Ventral", "NSC3")
names(avg_exp_by_cluster) <- c("Low Cerebellum", "Hippocampus/Cortex", "Cortical oRG", "G2/M", "Hypothalamus", "Cortical IPC", "Ventral", "oRG", "Hippocampus/Cortex", "Low Bias", "Thalamus/Cerebellum",  "Cerebellum", "Mostly G2/M", "Ventral", "NSC3")

# Save heatmaps and cluster assignments
#saveRDS(Atreyo_contacts_atlas_kmeans, file = "./Atreyo_neuron_kmeans_clusters.rds")
#Atreyo_contacts_atlas_kmeans <- readRDS(file ="./Atreyo_NSC_kmeans_clusters.rds")
#saveRDS(ann_heatmaps_by_cluster, file = "./Atreyo_neuron_contact_heatmaps.rds")
#ann_heatmaps_by_cluster <- readRDS(file = "./Atreyo_neuron_contact_heatmaps.rds")
```

Creating UMAP visualizations for embryonic and fetal neuronal atlas data

``` {r create_NSC_fetal_atlas_plots}
# Instantiate color maps for NSC subtypes and brain regions
fetal_atlas_colormap =  hue_pal()(29)
fetal_region_colormap = hue_pal()(8)

names(fetal_atlas_colormap) <- sort(unique(fetal_human_atlas_seurat_progs$cell_type))
names(fetal_region_colormap) <- sort(unique(fetal_human_atlas_seurat_progs$origin))

# Fetal atlas plot all cell types
DimPlot(fetal_human_atlas_seurat, group.by = "cell_type", pt.size = 1) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "bottom", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm')) + guides(color = guide_legend(override.aes = list(size = 6)))

# Fetal atlas plot all cell types
png(filename = "./panprogenitor_fetal_atlas_CXORG_highlight_UMAP.png", width = 575, height = 525)
DimPlot(fetal_human_atlas_seurat_progs, pt.size = 1, cells.highlight = WhichCells(fetal_human_atlas_seurat_progs, idents = "CX_ORG (G1/S)"), cols.highlight = fetal_region_colormap[['CX']]) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm')) + guides(color = guide_legend(override.aes = list(size = 6)))
dev.off()

# Fetal atlas plot regional origin
png(filename = "./panprogenitor_fetal_atlas_region_UMAP.png", width = 575, height = 525)
DimPlot(fetal_human_atlas_seurat_progs, group.by = "origin", pt.size = 1) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm'))
dev.off()

# Fetal atlas plot regional origin plus oRG and IPC
fetal_human_atlas_seurat_progs$origin_plus <- fetal_human_atlas_seurat_progs$origin
levels(fetal_human_atlas_seurat_progs$origin_plus) <- c(levels(fetal_human_atlas_seurat_progs$origin_plus), "CX_ORG", "CX_IPC")
fetal_human_atlas_seurat_progs$origin_plus[startsWith(fetal_human_atlas_seurat_progs$cell_type, "CX_ORG")] <- "CX_ORG"
fetal_human_atlas_seurat_progs$origin_plus[startsWith(fetal_human_atlas_seurat_progs$cell_type, "CX_IPC")] <- "CX_IPC"
fetal_region_colormap[['CX_ORG']] <- "#FDD017"
fetal_region_colormap[['CX_IPC']] <- "#BBC2CC"
fetal_human_atlas_seurat_progs$origin_plus <- relevel(fetal_human_atlas_seurat_progs$origin_plus, ref = "CX_IPC") %>% relevel(fetal_human_atlas_seurat_progs$origin_plus, ref = "CX_ORG")
png(filename = "./panprogenitor_fetal_atlas_region_plus_UMAP.png", width = 575, height = 525)
DimPlot(fetal_human_atlas_seurat_progs, group.by = "origin_plus", pt.size = 1, order = levels(fetal_human_atlas_seurat_progs$origin_plus)) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm')) + scale_color_manual(values = fetal_region_colormap)
dev.off()

# Density plots on panprogenitor atlas UMAP
png(filename = "./NSC_density_AUTS2.png", width = 575, height = 525)
plot_density(fetal_human_atlas_seurat, features = c("AUTS2"), pal = "magma") + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm'))
dev.off()
```

Create supplementary table of NSC HAR contact cluster assignments

``` {r create_supp_NSC_table}
# Create supplementary table of cluster assignments for genes
avg_exp_by_cluster <- lapply(1:15, function(x){NSC_contacts_atlas_avg_exp[neuron_contacts_atlas_kmeans$cluster == x,]})
avg_exp_by_cluster <- lapply(avg_exp_by_cluster, function(x) {as.data.frame(x) %>% mutate(gene = rownames(x))})
avg_exp_by_cluster <- lapply(1:15, function(x) {avg_exp_by_cluster[[x]] %>% 
    mutate(cluster = rep.int(x, dim(avg_exp_by_cluster[[x]])[1]))})
cluster_gene_assignment <- do.call("rbind", avg_exp_by_cluster) %>% select(gene, cluster)
cluster_name_list <- names(avg_exp_by_cluster)
names(cluster_name_list) <- 1:15
cluster_gene_assignment$cluster_name <- cluster_name_list[cluster_gene_assignment$cluster]

write.table(cluster_gene_assignment, 
            file = "~/Dropbox/Noonan Lab/workspace/Atreyo_HAR_contact_expression/NSC_cluster_assignment_supp_table.tsv",
            sep = "\t", row.names = FALSE, col.names = TRUE)
cluster_gene_assignment <- read.table(file = "~/Dropbox/Noonan Lab/workspace/Atreyo_HAR_contact_expression/NSC_cluster_assignment_supp_table.tsv", sep = "\t", header = TRUE)
```

Generating heatmaps of clustered average scaled expression values for neuronal HAR contacts

``` {r load_h5ad}
# Load in fetal and embryonic human neuron atlas (Bhaduri et al. 2021, Eze et al. 2021)
fetal_human_atlas_seurat <- LoadH5Seurat(file = "fetal_human_atlas_seurat.h5Seurat")

# Load in HAR contacts from in vitro neurons 
neuron_HAR_targets <- read.table("./contact_lists/mapping_hum_neurons.txt")
colnames(neuron_HAR_targets) <- c("HAR", "target")
neuron_HAR_targets <- separate_rows(neuron_HAR_targets, target, sep=",")

# Scale data and extract average scaled expression values for all neuron HAR contacts
fetal_human_atlas_seurat <- ScaleData(fetal_human_atlas_seurat)
neuron_contacts_atlas_avg_exp <- AverageExpression(fetal_human_atlas_seurat,
                                                     features = unique(neuron_HAR_targets$target), slot = "scale.data")
neuron_contacts_atlas_avg_exp <- neuron_contacts_atlas_avg_exp$RNA

# Extract average normalized expression values for all neuron HAR contacts (for downstream ordering of heatmaps)
neuron_contacts_atlas_avg_norm_exp <- AverageExpression(fetal_human_atlas_seurat,
                                                     features = unique(neuron_HAR_targets$target), slot = "data")
neuron_contacts_atlas_avg_norm_exp <- neuron_contacts_atlas_avg_norm_exp$RNA

# Group the average scaled expression values of neuron HAR contacts into 15 clusters using kmeans
num_centers = 15
neuron_contacts_atlas_kmeans <- kmeans(neuron_contacts_atlas_avg_exp, centers = num_centers)

# Establish color scheme for eight brain regions
region_colors <- hue_pal()(8)
names(region_colors) <- sort(unique(fetal_human_atlas_seurat$origin))

# Extract regional origin from atlas metadata and match to regional color
regional_origin <- fetal_human_atlas_seurat@meta.data %>% select(cell_type, origin) %>% group_by(cell_type) %>% summarize(regional_origin = names(which.max(table(origin)))) %>% mutate(regional_color = region_colors[regional_origin])

# Create regional origin annotation for heatmaps
region_annotation <- HeatmapAnnotation(region = regional_origin$regional_origin, col = list("region" = regional_origin$regional_color), annotation_label = "Regional Origin", annotation_name_gp = gpar(fontsize = 8), show_legend = FALSE, show_annotation_name = FALSE)

# Iterate through each cluster, creating a heatmap visualizing average scaled expression for genes within that cluster, ordered by the greatest normalized expression of each gene across neuronal subtypes
ann_heatmaps_by_cluster <- lapply(1:num_centers, function(x) {
draw(Heatmap(neuron_contacts_atlas_avg_exp[neuron_contacts_atlas_kmeans$cluster == x,][names(sort.int(apply(neuron_contacts_atlas_avg_norm_exp[rownames(neuron_contacts_atlas_avg_exp[neuron_contacts_atlas_kmeans$cluster == x,]),], 1, max), decreasing = TRUE)),],
                                          name = glue::glue("Cluster { x }"),
                                          bottom_annotation = region_annotation,
                                          row_names_gp = gpar(fontsize = 10),
                                          show_column_dend = TRUE,
                                          show_row_dend = FALSE,
                                          cluster_columns = TRUE,
                                          cluster_rows = FALSE,
                                          show_heatmap_legend = FALSE,
                                          column_names_rot = 70), padding = unit(c(2, 10, 2, 2), "mm"))})

# Annotation of heatmaps and average expression by neuronal subtype or regional bias
names(ann_heatmaps_by_cluster) <- c("Thalamic LGN Interneuron", "Hippocampus/Cortex/Cerebellum", "Hypothalamus/Hippocampus/Thalamus", "Ventral", "Low Enrichment", "Cerebellum", "Thalamic LGN Interneuron", "Hypothalamus/Hippocampus/Thalamus", "Hippocampus HCN3", "Hippocampus", "Low Interneuron", "Hypothalamic Glutamatergic Neuron", "Thalamus/Hypothalamus", "Hippocampal Dentate Gyrus", "Hippocampus/Cerebellum")
names(avg_exp_by_cluster) <- c("Thalamic LGN Interneuron", "Hippocampus/Cortex/Cerebellum", "Hypothalamus/Hippocampus/Thalamus", "Ventral", "Low Enrichment", "Cerebellum", "Thalamic LGN Interneuron", "Hypothalamus/Hippocampus/Thalamus", "Hippocampus HCN3", "Hippocampus", "Low Interneuron", "Hypothalamic Glutamatergic Neuron", "Thalamus/Hypothalamus", "Hippocampal Dentate Gyrus", "Hippocampus/Cerebellum")

# Save heatmaps and cluster assignments
#saveRDS(ann_heatmaps_by_cluster, file = "./HAR_neuron_contact_heatmaps.rds")
#ann_heatmaps_by_cluster <- readRDS(file = "~/Dropbox/Noonan Lab/workspace/human_chimp_organoid_project/Eze_et_al_2021_analysis/New_Eze_analysis/atlas_h5ad_objects/Atreyo_contact_lists/Atreyo_neuron_contact_heatmaps.rds")
#saveRDS(Atreyo_contacts_atlas_kmeans, file = "./Atreyo_neuron_kmeans_clusters.rds")
#Atreyo_contacts_atlas_kmeans <- readRDS(file = "./Atreyo_neuron_kmeans_clusters.rds")

```

Creating UMAP visualizations for embryonic and fetal neuronal atlas data

``` {r create_fetal_atlas_plots}
# Instantiate color maps for neuronal subtypes and brain regions
fetal_atlas_colormap =  hue_pal()(35)
fetal_region_colormap = hue_pal()(8)

names(fetal_atlas_colormap) <- sort(unique(fetal_human_atlas_seurat$cell_type))

# Fetal atlas plot all cell types
png(filename = "./panneuronal_fetal_atlas_UMAP.png", width = 850, height = 1000)
DimPlot(fetal_human_atlas_seurat, group.by = "cell_type", pt.size = 1) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "bottom", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm')) + guides(color = guide_legend(override.aes = list(size = 6)))
dev.off()

# Fetal atlas plot regional origin
fetal_human_atlas_seurat$origin_plus <- fetal_human_atlas_seurat$origin
levels(fetal_human_atlas_seurat$origin_plus) <- c(levels(fetal_human_atlas_seurat$origin_plus), "TH_LGNIN")
fetal_human_atlas_seurat$origin_plus[fetal_human_atlas_seurat$cell_type == "TH_LGNIN"] <- "TH_LGNIN"
fetal_human_atlas_seurat$origin_plus <- relevel(fetal_human_atlas_seurat$origin_plus, ref = "TH_LGNIN")
fetal_region_colormap[['TH_LGNIN']] <- "#FDD017"
png(filename = "./panneuronal_fetal_atlas_region_plus_UMAP.png", width = 575, height = 525)
DimPlot(fetal_human_atlas_seurat, group.by = "origin_plus", pt.size = 1, order = TRUE) + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm')) + scale_color_manual(values = fetal_region_colormap)
dev.off()

# Density plots on panneuronal atlas UMAP
png(filename = "./raw_plots/neuron_density_CDH4.png", width = 575, height = 525)
plot_density(fetal_human_atlas_seurat, features = c("PLAGL1"), pal = "magma") + th_cell_type_dimplot_theme + theme(axis.line = element_blank(), axis.ticks = element_blank(), legend.position = "none", plot.title = element_blank(), legend.spacing.x = unit(5, 'mm'))
dev.off()
```

Create supplementary table of neuron HAR contact cluster assignments

``` {r create_supp_neuron_table}
# Create supplementary table of cluster assignments for genes
avg_exp_by_cluster <- lapply(1:15, function(x){neuron_contacts_atlas_avg_exp[neuron_contacts_atlas_kmeans$cluster == x,]})
avg_exp_by_cluster <- lapply(avg_exp_by_cluster, function(x) {as.data.frame(x) %>% mutate(gene = rownames(x))})
avg_exp_by_cluster <- lapply(1:15, function(x) {avg_exp_by_cluster[[x]] %>% 
    mutate(cluster = rep.int(x, dim(avg_exp_by_cluster[[x]])[1]))})
cluster_gene_assignment <- do.call("rbind", avg_exp_by_cluster) %>% select(gene, cluster)
cluster_name_list <- names(avg_exp_by_cluster)
names(cluster_name_list) <- 1:15
cluster_gene_assignment$cluster_name <- cluster_name_list[cluster_gene_assignment$cluster]

write.table(cluster_gene_assignment, 
            file = "~/Dropbox/Noonan Lab/workspace/Atreyo_HAR_contact_expression/neuron_cluster_assignment_supp_table.tsv", 
            sep = "\t", row.names = FALSE, col.names = TRUE)
cluster_gene_assignment_neuron <- read.table(file = "~/Dropbox/Noonan Lab/workspace/Atreyo_HAR_contact_expression/neuron_cluster_assignment_supp_table.tsv", sep = "\t", header = TRUE)
```

DEG HAR contact analysis

Create heatmap for oRG-biased, human-biased DEG NSC HAR contacts

``` {r Atreyo_heatmaps}
# Load in DEG HAR contacts and filter to only those with human-biased expression
DE_genes_HAR_targets <- read.table("./contact_lists/DEG_table_corrected_apeglm.txt", row.names = NULL, header = TRUE)
#colnames(DE_genes_HAR_targets) <- c("Gene", "DHM", "DEG")
DE_genes_HAR_targets <- DE_genes_HAR_targets %>% filter(log2FoldChange > 0 & gene_name %in% NSC_HAR_targets$target) %>% select(gene_name) %>% unique()

# Scale data and extract average scaled expression values for all human-biased DEG NSC HAR contacts
fetal_human_atlas_seurat_progs <- ScaleData(fetal_human_atlas_seurat_progs)
NSC_DEG_contacts_atlas_avg_exp <- AverageExpression(fetal_human_atlas_seurat_progs,
                                                     features = DE_genes_HAR_targets$gene_name, slot = "scale.data")
NSC_DEG_contacts_atlas_avg_exp <- NSC_DEG_contacts_atlas_avg_exp$RNA

# Extract average normalized expression values for all DEG NSC HAR contacts (for downstream ordering of heatmaps)
NSC_DEG_contacts_atlas_avg_norm_exp <- AverageExpression(fetal_human_atlas_seurat_progs,
                                                     features = unique(DE_genes_HAR_targets$gene_name), slot = "data")
NSC_DEG_contacts_atlas_avg_norm_exp <- NSC_DEG_contacts_atlas_avg_norm_exp$RNA

# Rename NSC subtypes for downstream visualization
colnames(NSC_DEG_contacts_atlas_avg_exp) <- c("CB_NSC2 (G1/S)", "CB_NSC2 (G2/M)", "CB_NSC (G1/S)", "CB_NSC (G2/M)", "CB_NSC3 (G1/S)", "CGE_NSC (G1/S)", "CGE_NSC (G2/M)", "CX_IPC (G1/S)", "CX_IPC (G2/M)", "CX_ORG (G1/S)", "CX_RG (G1/S)", "CX_RG (G2/M)", "HC_NSC2 (G1/S)", "HC_NSC2 (G2/M)", "HC_NSC (G1/S)", "HC_NSC (G2/M)", "HC_NSC3 (G1/S)", "HT_NSC2 (G1/S)", "HT_NSC2 (G2/M)", "HT_NSC (G1/S)", "HT_NSC (G2/M)", "HT_NSC3 (G1/S)", "HT_NSC3 (G2/M)", "LGE_NSC (G1/S)", "LGE_NSC (G2/M)", "MGE_NSC (G1/S)", "MGE_NSC (G2/M)", "TH_NSC (G1/S)", "TH_NSC (G2/M)")
colnames(NSC_DEG_contacts_atlas_avg_exp) <- str_split_fixed(colnames(NSC_DEG_contacts_atlas_avg_exp), " ", 2)[,1]
colnames(NSC_DEG_contacts_atlas_avg_norm_exp) <- c("CB_NSC2 (G1/S)", "CB_NSC2 (G2/M)", "CB_NSC (G1/S)", "CB_NSC (G2/M)", "CB_NSC3 (G1/S)", "CGE_NSC (G1/S)", "CGE_NSC (G2/M)", "CX_IPC (G1/S)", "CX_IPC (G2/M)", "CX_ORG (G1/S)", "CX_RG (G1/S)", "CX_RG (G2/M)", "HC_NSC2 (G1/S)", "HC_NSC2 (G2/M)", "HC_NSC (G1/S)", "HC_NSC (G2/M)", "HC_NSC3 (G1/S)", "HT_NSC2 (G1/S)", "HT_NSC2 (G2/M)", "HT_NSC (G1/S)", "HT_NSC (G2/M)", "HT_NSC3 (G1/S)", "HT_NSC3 (G2/M)", "LGE_NSC (G1/S)", "LGE_NSC (G2/M)", "MGE_NSC (G1/S)", "MGE_NSC (G2/M)", "TH_NSC (G1/S)", "TH_NSC (G2/M)")
colnames(NSC_DEG_contacts_atlas_avg_norm_exp) <- str_split_fixed(colnames(NSC_DEG_contacts_atlas_avg_norm_exp), " ", 2)[,1]

# Create heatmap of all 88 human-biased DEG NSC HAR contacts
Heatmap(as.matrix(NSC_DEG_contacts_atlas_avg_exp),
        row_names_gp = gpar(fontsize = 10),
        cluster_columns = TRUE,
        bottom_annotation = region_annotation,
        top_annotation = cell_cycle_annotation,
        show_heatmap_legend = FALSE,
        column_names_rot = 70)

# Isolate oRG-/cortex-biased genes by partitioning 88 DEG NSC HAR contacts into two clusters (results in one cluster of 20 genes with oRG bias)
num_centers = 3
NSC_DEG_contacts_atlas_kmeans <- kmeans(NSC_DEG_contacts_atlas_avg_exp, centers = num_centers)

# Establish color scheme for eight brain regions and cell cycle phase
cell_cycle_colors <- c("light grey", "#386C0B")
names(cell_cycle_colors) <- c("(G1/S)", "(G2/M)")
region_colors <- hue_pal()(8)
names(region_colors) <- sort(unique(fetal_human_atlas_seurat_progs$origin))

# Extract regional origin and cell cycle from atlas metadata and associate with respective color map
regional_origin <- fetal_human_atlas_seurat_progs@meta.data %>% select(cell_type, origin) %>% group_by(cell_type) %>% summarize(regional_origin = names(which.max(table(origin)))) %>% mutate(regional_color = region_colors[regional_origin])
#regional_origin$cell_type <- colnames(NSC_DEG_contacts_atlas_avg_exp)
regional_origin <- regional_origin %>% mutate(cell_cycle = str_split_fixed(cell_type, " ", 2)[,2])
regional_origin$cell_cycle[5] <- "(G1/S)"
regional_origin$cell_cycle[17] <- "(G1/S)"
regional_origin$cell_cycle_colors <- cell_cycle_colors[regional_origin$cell_cycle]

# Create regional origin and cell cycle phase annotations for heatmaps
region_annotation <- HeatmapAnnotation(region = regional_origin$regional_origin, col = list("region" = regional_origin$regional_color), annotation_label = "Regional Origin", annotation_name_gp = gpar(fontsize = 8), show_legend = FALSE)
cell_cycle_annotation <- HeatmapAnnotation(cell_cycle = regional_origin$cell_cycle, col = list("cell_cycle" = regional_origin$cell_cycle_colors), annotation_label = "Cell Cycle Phase", annotation_name_gp = gpar(fontsize = 8), show_legend = FALSE)

# Create two heatmaps from the human-biased DEG NSC HAR contacts
ann_heatmaps_by_cluster <- lapply(1:num_centers, function(x) {
draw(Heatmap(as.matrix(NSC_DEG_contacts_atlas_avg_exp[NSC_DEG_contacts_atlas_kmeans$cluster == x,][names(sort.int(apply(NSC_DEG_contacts_atlas_avg_norm_exp[rownames(NSC_DEG_contacts_atlas_avg_exp[NSC_DEG_contacts_atlas_kmeans$cluster == x,]),], 1, max), decreasing = TRUE)),]),
                                          name = glue::glue("Cluster { x }"),
                                          bottom_annotation = cell_cycle_annotation,
                                          top_annotation = region_annotation,
                                          row_names_gp = gpar(fontsize = 10),
                                          show_column_dend = TRUE,
                                          show_row_dend = FALSE,
                                          cluster_columns = TRUE,
                                          cluster_rows = FALSE,
                                          show_heatmap_legend = FALSE,
                                          column_names_rot = 70), padding = unit(c(2, 10, 2, 2), "mm"))})
```



